The Array of Long-Baseline Antennas for Taking Radio Observations from the Sub-Antarctic

### 2020-08-13 albatros_paper.tex
HCC embiggened the last figure
HCC updated baseband/direct autospectrum comparison figure and text

The primary requirements that drive the design for a low-frequency
imaging experiment are 1) desired resolution, 2) low RFI, and 3) quiet
ionospheric conditions.  As a benchmark, an interferometer operating
at 30~MHz requires baseline lengths of $\sim0.5$--1~km 
to achieve a resolution of $\sim 1$\degree.  This length scales
inversely with frequency, and therefore $\sim 10$-km lengths are
required at few-MHz frequencies to improve substantially upon the resolutions
achieved to date.  The experiment installation site must be remote to
keep RFI to a minimum, and polar or near-polar latitiudes generally
have lower ionospheric plasma frequency cutoffs relative to other
locations on Earth.  \autoref{Fig:iri} illustrates predictions from
the International Reference Ionosphere
model~\citep{2018AdRS...16....1B} for the minimum plasma frequency
during the last solar minimum, which started in approximately 2007.
Three locations are shown: Marion Island, the focus of this work;
Dome~C in Antarctica, which is another isolated location used for
astronomical observations; and Hobart, Tasmania, where Reber performed
his $\sim2$~MHz observations.  The model predictions illustrate that
the ionospheric plasma cutoff frequency at Marion Island may drop as
low as $\sim1.5$~MHz during solar minima.  Since we are currently
experiencing another solar minimum~\citep{2018NatCo...9.5209B}, the
timing is opportune for new low-frequency observations.

@@

\begin{figure}
  \begin{center} \includegraphics[width=1.0\linewidth]{Figures/albatros_2elem_schematic/albatros_2elem_schematic.pdf}
    \caption{Two-element \albatros\ pathfinder block diagram.  Signals
      from two dual-polarization LWA antennas are amplified by
      front-end active baluns~\citep{2012PASP..124.1090H}.  The
      antennas are connected via 100-m coaxial cables to the back-end
      readout electronics, which are housed in a Faraday cage denoted
      by the dashed box.  Each of the four antenna outputs is passed
      to a second-stage electronics chain consisting of filters and
      further amplfication.  The signals are digitized at 250~Msamp/s
      by a SNAP board, which includes an on-board FPGA that computes
      auto- and cross-spectra from and between the four inputs.  A
      Raspberry Pi controls the SNAP board and saves the data.}
    \label{Fig:albatros2_schem}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{Figures/albatros_single_schematic/albatros_single_schematic.pdf}
    \caption{Single-antenna autonomous station block diagram.  A
      dual-polarization LWA antenna, equipped with a front-end active
      balun, connects via 50-m coaxial cables to the back-end readout
      electronics, housed in a Faraday cage denoted by the dashed box.
      Each of the two antenna signals is passed to a second-stage
      electronics chain consisting of filters and further
      amplfication.  The signals are digitized at 250~Msamp/s by a
      SNAP board, which includes an on-board FPGA that computes
      channelized baseband and spectra from both inputs.  A Raspberry
      Pi controls the SNAP board and receives the baseband data and
      spectra.  The baseband data are saved to external hard drives.
      The system is powered by solar panels that charge a 24-V battery
      bank.}
    \label{Fig:albatros1_schem}
  \end{center}
\end{figure}

@@

The autonomous station is powered by two series-connected, 12~V, deep
cycle, lead acid batteries, charged by an array of nine SunPower
SPE-E-Flex-110 solar panels. These panels each have a standard test
capacity of 110~W. Although only $\sim$50~W is required to run the
station, the $\sim$1~kW charging capacity is intentionally oversized
to account for the frequent overcast conditions on Marion.  The
generous power margin allows the charge level to recover quickly on a
short sunny day in winter in the event of charge loss over several
consecutive overcast days.  The nine solar panels are distributed
between three custom designed structures built from aluminum
extrusion, with rigid metalized plastic panels backing the
semi-flexible solar cells. The structures are oriented due north and
are designed to incline the solar panels at a relatively steep angle
to maximize peformance under winter conditions, when sunlight hours
are at their minimum. The solar panel mounts have been designed to
withstand frequent gale-force winds on Marion. While the mounting
structures themselves appear to be strong enough, the greatest
challenge in deployment has been to find adequate anchoring in the
volcanic ground.  Each group of three panels is wired in series, and
the three series strings are connected in parallel to a Victron
BlueSolar MPPT 50$\vert$35 charge controller, which optimizes power
transfer from the solar array when charging is required, and also
monitors charge level, reducing output current when the battery bank
is fully charged.

@@

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.9\linewidth]{Figures/albatros2_waterfalls.jpg}
    \caption{Spectra from a co-aligned polarization pair from the
      two-element \albatros\ pathfinder, shown as a function of
      frequency and time.  The auto- and cross-spectra are shown in
      the top and bottom rows, respectively.  The spectrum magnitudes
      are in uncalibrated ADC units on a logarithmic scale, and the
      cross-spectrum is about two orders of magnitude fainter than the
      autospectra.  Repeatable structure from the Galaxy is visible
      over the three-day time scale in all plots.}
    \label{Fig:albatros2_waterfalls}
  \end{center}
\end{figure}

@@

We have been unable to retrieve data from the single autonomous
station pathfinder on Marion because of COVID-19 and limited bandwidth
to the island, although we have confirmed that the solar power system
has successfully met the power demands of the antenna station.  To
demonstrate baseband writing capability in the absence of Marion data,
we instead present a short observation that was taken by a separate
\albatros\ pathfinder station that operated from the McGill Arctic
Research Station (MARS; \ang{79;26;}N, \ang{90;46;}W) on Axel Heiberg
Island, Nunavut, during July~2019.  The RF signal chain of the MARS
installation had the same architecture as that shown in
\autoref{Fig:albatros1_schem}, but the power was supplied by manually
charged batteries, rather than a solar power system.  The MARS RFI
environment is somewhat noisier than that of Marion Island, and
because the data were taken during the height of Arctic summer, the
ionospheric plasma cutoff frequency is higher than the typical cutoff
reflected in the Marion data.  Despite the differences in the power
configuration and RFI environment, the MARS installation is suitable
for illustrating baseband data writing.

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.8\linewidth]{Figures/baseband_direct_auto_compare.jpg}
    \caption{\attention{Placeholder figure} Autospectra for one
      polarization from an \albatros\ pathfinder at MARS (see text for
      details).  The top row shows spectra that are accumulated
      directly by the SNAP board, and the middle row shows spectra
      computed from baseband data with 4-bit quantization.  The two
      frequency windows, 5.3--12.6~MHz and 12.6--20.0~MHz, are
      recorded at two separate times, and all waterfall plots are
      shown with 61-kHz resolution.  The bottom row compares the mean
      autospectra, with finer, 14.6-kHz resolution of
      the baseband spectra.}
    \label{Fig:baseband_direct_auto}
  \end{center}
\end{figure}

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.8\linewidth]{Figures/baseband_direct_phase_compare.png}
    \caption{Phases from the cross-correlation between two orthogonal
      polarizations from an \albatros\ pathfinder at MARS (see text
      for details).  The top row shows cross-spectrum phases that are
      accumulated directly by the SNAP board, and the bottom row shows
      phases computed from baseband data with 4-bit quantization.  The
      two frequency windows, 5.3--12.6~MHz and 12.6--20.0~MHz, are
      recorded at two separate times, and all waterfall plots are
      shown with 61-kHz resolution.}
    \label{Fig:baseband_direct_phase}
  \end{center}
\end{figure}

\autoref{Fig:baseband_direct_auto} shows autospectra for one
polarization of the \albatros\ pathfinder installed at MARS, comparing
the accumulated spectra that are directly computed by the SNAP board
against spectra that are calculated from baseband data with 4-bit
quantization.  (Autospectrum information is not preserved with 1-bit
quantization.)  Baseband data were recorded over two frequency
windows, 5.3--12.6~MHz and 12.6--20.0~MHz, at separate times, and the
corresponding accumulated spectra from the SNAP board are shown for
the same frequency and time ranges.  All of the waterfall plots have
the same 61-kHz frequency resolution, set by the resolution of the
spectra computed by the SNAP board.  In the overplotted mean spectra,
the baseband data are shown with finer, 14.6-kHz resolution to
illustrate the additional information that is not available in the
accumulated spectra.  For these mean spectrum plots, the baseband data
are processed by
inverting\footnote{\url{https://github.com/jrs65/pfb-inverse}} the
polyphase filter bank to recover time-domain data, and the
forward-direction polyphase filter bank is then applied to
concatenated timestreams of an appropriate total length to set the
desired frequency resolution.  \autoref{Fig:baseband_direct_phase}
shows waterfall plots of phases computed from the cross-correlation
between the two orthogonal polarizations of the \albatros\ pathfinder
at MARS, comparing the directly accumulated and baseband data.  For
both the autospectra and phases, the baseband data qualitatively agree
with the data directly accumulated by the SNAP board, up to small
differences caused by the 4-bit quantization.

@@

With the proof of concept demonstrated from the pathfinder instruments
presented here, we plan to improve the design of the future
\albatros\ stations that will be installed at the coastal huts on
Marion.  The LWA antenna and FEE are currently not optimized for the
lowest \albatros\ observing frequencies, and we are investigating
possible modifications to the antenna and FEE designs to improve the
low frequency response, as well as building in calibration circuitry.
Because the future \albatros\ stations will be located farther from
the base and will be more difficult to access on a regular basis, each
station will require larger total hard drive volume to store baseband
data over extended periods of time.  We are developing a custom
low-power hard drive bank with $>100$~TB total storage capacity that
will employ a USB multiplexer to select and power only one hard drive
at a time.  The autonomous station pathfinder presented here uses
solar panels to charge the batteries, and we are investigating the use
of small wind turbines as a possible alternative for future stations.

In addition to Marion Island, we are also planning on installing a
second \albatros\ array at MARS in the high Arctic.  As described in
\S\ref{s:obs}, a pathfinder antenna was installed in July 2019 and
observed for about three weeks to assess the RFI environment and
ionospheric conditions.  When fully operational, the Marion and MARS
\albatros\ installations will provide new views of the low-frequency
sky across both hemispheres.

>>>>

The primary requirements that drive the design for a low-frequency
imaging experiment are 1) desired resolution, 2) low RFI, and 3) quiet
ionospheric conditions.  As a benchmark, an interferometer operating
at 30~MHz requires baseline lengths of $\sim0.5$--1~km 
to achieve a resolution of $\sim 1$\degree.  This length scales
inversely with frequency, and therefore $\sim 10$-km lengths are
required at few-MHz frequencies to improve substantially upon the resolutions
achieved to date.  The experiment installation site must be remote to
keep RFI to a minimum, and polar or near-polar latitudes generally
have lower ionospheric plasma frequency cutoffs relative to other
locations on Earth.  \autoref{Fig:iri} illustrates predictions from
the International Reference Ionosphere
model~\citep{2018AdRS...16....1B} for the minimum plasma frequency
during the last solar minimum, which started in approximately 2007.
Three locations are shown: Marion Island, the focus of this work;
Dome~C in Antarctica, which is another isolated location used for
astronomical observations; and Hobart, Tasmania, where Reber performed
his $\sim2$~MHz observations.  The model predictions illustrate that
the ionospheric plasma cutoff frequency at Marion Island may drop as
low as $\sim1.5$~MHz during solar minima.  Since we are currently
experiencing another solar minimum~\citep{2018NatCo...9.5209B}, the
timing is opportune for new low-frequency observations.

@@

\begin{figure}
  \begin{center} \includegraphics[width=1.0\linewidth]{Figures/albatros_2elem_schematic/albatros_2elem_schematic.pdf}
    \caption{Two-element \albatros\ pathfinder block diagram.  Signals
      from two dual-polarization LWA antennas are amplified by
      front-end active baluns~\citep{2012PASP..124.1090H}.  The
      antennas are connected via 100-m coaxial cables to the back-end
      readout electronics, which are housed in a Faraday cage denoted
      by the dashed box.  Each of the four antenna outputs is passed
      to a second-stage electronics chain consisting of filters and
      further amplification.  The signals are digitized at 250~Msamp/s
      by a SNAP board, which includes an on-board FPGA that computes
      auto- and cross-spectra from and between the four inputs.  A
      Raspberry Pi controls the SNAP board and saves the data.}
    \label{Fig:albatros2_schem}
  \end{center}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{Figures/albatros_single_schematic/albatros_single_schematic.pdf}
    \caption{Single-antenna autonomous station block diagram.  A
      dual-polarization LWA antenna, equipped with a front-end active
      balun, connects via 50-m coaxial cables to the back-end readout
      electronics, housed in a Faraday cage denoted by the dashed box.
      Each of the two antenna signals is passed to a second-stage
      electronics chain consisting of filters and further
      amplification.  The signals are digitized at 250~Msamp/s by a
      SNAP board, which includes an on-board FPGA that computes
      channelized baseband and spectra from both inputs.  A Raspberry
      Pi controls the SNAP board and receives the baseband data and
      spectra.  The baseband data are saved to external hard drives.
      The system is powered by solar panels that charge a 24-V battery
      bank.}
    \label{Fig:albatros1_schem}
  \end{center}
\end{figure}

@@

The autonomous station is powered by two series-connected, 12~V, deep
cycle, lead acid batteries, charged by an array of nine SunPower
SPE-E-Flex-110 solar panels. These panels each have a standard test
capacity of 110~W. Although only $\sim$50~W is required to run the
station, the $\sim$1~kW charging capacity is intentionally oversized
to account for the frequent overcast conditions on Marion.  The
generous power margin allows the charge level to recover quickly on a
short sunny day in winter in the event of charge loss over several
consecutive overcast days.  The nine solar panels are distributed
between three custom designed structures built from aluminum
extrusion, with rigid metalized plastic panels backing the
semi-flexible solar cells. The structures are oriented due north and
are designed to incline the solar panels at a relatively steep angle
to maximize performance under winter conditions, when sunlight hours
are at their minimum. The solar panel mounts have been designed to
withstand frequent gale-force winds on Marion. While the mounting
structures themselves appear to be strong enough, the greatest
challenge in deployment has been to find adequate anchoring in the
volcanic ground.  Each group of three panels is wired in series, and
the three series strings are connected in parallel to a Victron
BlueSolar MPPT 50$\vert$35 charge controller, which optimizes power
transfer from the solar array when charging is required, and also
monitors charge level, reducing output current when the battery bank
is fully charged.

@@

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=\linewidth]{Figures/albatros2_waterfalls.jpg}
    \caption{Spectra from a co-aligned polarization pair from the
      two-element \albatros\ pathfinder, shown as a function of
      frequency and time.  The auto- and cross-spectra are shown in
      the top and bottom rows, respectively.  The spectrum magnitudes
      are in uncalibrated ADC units on a logarithmic scale, and the
      cross-spectrum is about two orders of magnitude fainter than the
      autospectra.  Repeatable structure from the Galaxy is visible
      over the three-day time scale in all plots.}
    \label{Fig:albatros2_waterfalls}
  \end{center}
\end{figure}

@@

We have been unable to retrieve data from the single autonomous
station pathfinder on Marion because of COVID-19 restrictions and
limited bandwidth to the island, although we have confirmed that the
solar power system has successfully met the power demands of the
antenna station.  To demonstrate baseband writing capability in the
absence of Marion data, we instead present a short observation that
was taken by a separate, single-element \albatros\ pathfinder station
that operated from the McGill Arctic Research Station (MARS;
\ang{79;26;}N, \ang{90;46;}W) on Axel Heiberg Island, Nunavut, during
July~2019.  The RF signal chain of the MARS installation had the same
architecture as that shown in \autoref{Fig:albatros1_schem}, but the
power was supplied by manually charged batteries, rather than a solar
power system.  The MARS RFI environment is somewhat noisier than that
of Marion Island, and because the data were taken during the height of
Arctic summer, the ionospheric plasma cutoff frequency is higher than
the typical cutoff reflected in the Marion data.  Despite the
differences in the power configuration and RFI environment, the MARS
installation is suitable for illustrating baseband data writing.

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=\linewidth]{Figures/baseband_direct_auto_compare.png}
    \caption{Autospectra for one polarization from an
      \albatros\ pathfinder at MARS (see text for details).  The top
      row shows spectra that are accumulated directly by the SNAP
      board, and the middle row shows spectra computed from baseband
      data with 4-bit quantization.  The two frequency windows,
      5.3--12.6~MHz and 12.6--20.0~MHz, are recorded at two separate
      times, and all plots are shown with 61-kHz resolution.  The
      waterfall plots are shown with a logarithmic color scale to
      highlight the qualitative agreement in spectral features between
      the directly-accumulated and baseband spectra.  The bottom row
      compares the mean autospectra, which agree well when the
      baseband data are not saturated by bright RFI emission.}
    \label{Fig:baseband_direct_auto}
  \end{center}
\end{figure}

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=\linewidth]{Figures/baseband_direct_phase_compare.png}
    \caption{Phases from the cross-correlation between two orthogonal
      polarizations from an \albatros\ pathfinder at MARS (see text
      for details).  The top row shows cross-spectrum phases that are
      accumulated directly by the SNAP board, and the bottom row shows
      phases computed from baseband data with 4-bit quantization.  The
      two frequency windows, 5.3--12.6~MHz and 12.6--20.0~MHz, are
      recorded at two separate times, and all waterfall plots are
      shown with 61-kHz resolution.}
    \label{Fig:baseband_direct_phase}
  \end{center}
\end{figure}

\autoref{Fig:baseband_direct_auto} shows autospectra for one
polarization of the \albatros\ pathfinder installed at MARS, comparing
the accumulated spectra that are directly computed by the SNAP board
against spectra that are calculated from baseband data with 4-bit
quantization.  (Autospectrum information is not preserved with 1-bit
quantization.)  Baseband data were recorded over two frequency
windows, 5.3--12.6~MHz and 12.6--20.0~MHz, at separate times, and the
corresponding accumulated spectra from the SNAP board are shown for
the same frequency and time ranges.  All of the waterfall plots have
the same 61-kHz frequency resolution, set by the resolution of the
spectra computed by the SNAP board, and are normalized so that the
maximum is unity.  The waterfall plots are shown with a logarithmic
scale to highlight the qualitative agreement between the
directly-accumulated and baseband spectra for bright and faint
spectral features.  The bottom panels show the time-averaged spectra
on a linear scale to illustrate differences in bright RFI features
that saturate the the baseband data.  The fraction of baseband data
with values at the 4-bit extrema are overplotted with spectra, and
there is agreement between the directly-accumulated and baseband
spectra except when saturation is present.

% In the overplotted mean spectra, the baseband data are shown with
% finer, 14.6-kHz resolution to illustrate the additional information
% that is not available in the accumulated spectra.  For these mean
% spectrum plots, the baseband data are processed by
% inverting\footnote{\url{https://github.com/jrs65/pfb-inverse}} the
% polyphase filter bank to recover time-domain data, and the
% forward-direction polyphase filter bank is then applied to
% concatenated timestreams of an appropriate total length to set the
% desired frequency resolution.  \autoref{Fig:baseband_direct_phase}
% shows waterfall plots of phases computed from the cross-correlation
% between the two orthogonal polarizations of the \albatros\ pathfinder
% at MARS, comparing the directly accumulated and baseband data.  For
% both the autospectra and phases, the baseband data qualitatively agree
% with the data directly accumulated by the SNAP board, up to small
% differences caused by the 4-bit quantization.

@@

With the proof of concept demonstrated from the pathfinder instruments
presented here, we plan to improve the design of the future
\albatros\ stations that will be installed at the coastal huts on
Marion.  The LWA antenna and FEE are currently not optimized for the
lowest \albatros\ observing frequencies, and we are investigating
possible modifications to the antenna and FEE designs to improve the
low frequency response, as well as building in calibration circuitry.
Because the future \albatros\ stations will be located farther from
the base and will be more difficult to access on a regular basis, each
station will require larger total hard drive volume to store baseband
data over extended periods of time.  We are developing a custom
low-power hard drive bank with $>100$~TB total storage capacity that
will employ a USB multiplexer to select and power only one hard drive
at a time.  The autonomous station pathfinder presented here uses
solar panels to charge the batteries, and we are investigating the use
of small wind turbines as a possible alternative for future stations.
We are also developing analysis tools to compute time-domain data from
the recorded channelized baseband by inverting the polyphase filter
bank while minimizing artifacts from quantization and saturation.

In addition to Marion Island, we are also planning on installing a
second \albatros\ array at MARS in the high Arctic.  As described in
\S\ref{s:obs}, a single pathfinder antenna was installed in July 2019
and observed for about three weeks to assess the RFI environment and
ionospheric conditions.  When fully operational, the Marion and MARS
\albatros\ installations will provide new views of the low-frequency
sky across both hemispheres.
